---
title: "EDA - Census Data"
author: "Ajay Tom"
date: "2023-03-20"
output: html_document
---

## <span style="font-size: 1.4em;">1. INTRODUCTION</span>

In today’s digital age, high-speed internet has become a necessity for people and businesses alike. With the advent of 5G, the latest and fastest wireless technology, internet speeds have reached a new level of lightning-fast connectivity. Our client, a leading telecom company called "ConnectX", has recognized the immense potential of this technology and aims to expand its 5G network across India. However, with a country as diverse and vast as India, rolling out 5G poses unique challenges. ConnectX understands the need to strategically plan and prioritize their 5G deployment in order to maximize their investments and provide the best possible service to their customers. In this report, we analyze India's census data to help ConnectX make data-driven decisions to optimize their 5G rollout strategy.

**PROBLEM STATEMENT:** How can we analyze India's census data to help ConnectX, a leading telecom company, make data-driven decisions and optimize their 5G rollout strategy?

## <span style="font-size: 1.15em;">Import Dataset</span>

```{r warning=FALSE , message=FALSE}
library(readxl)
census_data <- read_excel("D:\\GitHub\\EDA-census_data\\A-1_NO_OF_VILLAGES_TOWNS_HOUSEHOLDS_POPULATION_AND_AREA.xlsx")
```

## <span style="font-size: 1.4em;">2. HEADER CLEANING</span>

```{r warning=FALSE, message=FALSE}
library(DT)
head(census_data)
```
There are multiple header columns present in the dataset. These must be simplified and made one for better analysis.


## <span style="font-size: 1.15em;">Deleting 2nd row</span>
```{r}
census_data <- census_data[-2,]
head(census_data)
```
## <span style="font-size: 1.15em;">Rename Columns</span>
```{r warning=FALSE, message=FALSE}
library(dplyr)
names(census_data) #to see the names of the columns

  #changed 'Number of villages' column to NV_Inhabited
  census_data <- rename(census_data, NV_Inhabited = "Number of villages")
  #changed '...8' column to NV_Uninhabited
  census_data <- rename(census_data, NV_Uninhabited = "...8")
  
  #changed the population super column to 3 seperate columns 
  #P_Total, P_Males and P_Females
  census_data <- rename(census_data, P_Total = "Population")
  census_data <- rename(census_data, P_Males = "...12")
  census_data <- rename(census_data, P_Females = "...13")

names(census_data)
```
## <span style="font-size: 1.15em;">Delete the redundant 1st row</span>
```{r warning=FALSE, message=FALSE}
census_data <- census_data[-1,]
head(census_data)
```



## <span style="font-size: 1.4em;">3. DATA UNDERSTANDING</span>
```{r warning=FALSE, message=FALSE}
library(tidyverse)
str(census_data)
glimpse(census_data)
names(census_data)
```
## <span style="font-size: 1.4em;">4. MISSING VALUES</span>
```{r warning=FALSE, message=FALSE}
library(DataExplorer)
plot_missing(census_data)
```


## <span style="font-size: 1.4em;">5. JUNK VALUES</span>
Apart from the NAs, there are some junk values in  the observations which if not handled, would impact the analysis
```{r warning=FALSE, message=FALSE}
unique(census_data$`District Code`)
unique(census_data$`India/ State/ Union Territory/ District/ Sub-district`)
unique(census_data$`Total/\r\nRural/\r\nUrban`) 
```

## <span style="font-size: 1.15em;">Filtering out</span>
**1. District Code**

Filtering out non numeric values from **'census_data'** and saving it in a new dataframe **'census_data_filtered'**

```{r warning=FALSE, message=FALSE}
census_data_filtered <- census_data %>% 
                            filter(!grepl("[^0-9]",`District Code`)) 
```

Filtering out the rows having 1,2,3,4 and 5 as District Code values

```{r warning=FALSE, message=FALSE}
census_data_filtered <- census_data_filtered %>% 
                            filter(!`District Code` %in% c(1,2,3,4,5))
```

**2. Sub District Code**

Filtering out non numeric values from **Sub District Code** 
```{r warning=FALSE, message=FALSE}
census_data_filtered <- census_data_filtered %>% 
                            filter(!grepl("[^0-9]",`Sub District Code`))
```

**3. Number of households**

Replace 1e+05 with 100000 - R was only displaying 100000 as 1e+05. To display it as 100000:
```{r warning=FALSE, message=FALSE}
options(scipen = 999)
```

**4. Area\r\n (In sq. km)**

Filter out the junk files(excluding NAs)
```{r warning=FALSE, message=FALSE}
  census_data_filtered <- census_data_filtered %>% 
    filter(is.na(`Area\r\n (In sq. km)`) |
           (grepl("^\\d+\\.?\\d*$", `Area\r\n (In sq. km)`)))
```

**5. Remove special characters from Name column (eg. INDIA @& to INDIA)**

Filter out the junk values brought down the number of observations from 20020 to 19995
```{r warning=FALSE, message=FALSE}
  
head(census_data_filtered$Name) 
  census_data_filtered <- census_data_filtered %>% 
                            mutate(Name = str_replace_all(Name, "@&", ""))
  census_data_filtered <- census_data_filtered %>% 
                            mutate(Name = str_replace_all(Name, "\\$", ""))
  
head(census_data_filtered$Name)  
```


**Observations Reduced**

Filter out the junk values brought down the number of observations from 20020 to 19995
```{r warning=FALSE, message=FALSE}
  nrow(census_data)
  nrow(census_data_filtered)
```



## <span style="font-size: 1.4em;">6. TYPE CONVERSION</span>

Converting columns into appropriate data types
```{r warning=FALSE, message=FALSE}

census_data_filtered$`State  Code` <- as.factor(census_data_filtered$`State  Code`)
  census_data_filtered$`District Code` <- as.factor(census_data_filtered$`District Code`)
  census_data_filtered$`Sub District Code` <- as.factor(census_data_filtered$`Sub District Code`)
  census_data_filtered$`India/ State/ Union Territory/ District/ Sub-district` <- as.factor(census_data_filtered$`India/ State/ Union Territory/ District/ Sub-district`)
  census_data_filtered$`Name` <- as.character(census_data_filtered$`Name`)
  census_data_filtered$`Total/\r\nRural/\r\nUrban` <- as.factor(census_data_filtered$`Total/\r\nRural/\r\nUrban`)
  census_data_filtered$`NV_Inhabited` <- as.numeric(census_data_filtered$`NV_Inhabited`)
  census_data_filtered$`NV_Uninhabited` <- as.numeric(census_data_filtered$`NV_Uninhabited`)
  census_data_filtered$`Number of towns` <- as.numeric(census_data_filtered$`Number of towns`)
  census_data_filtered$`Number of households` <- as.numeric(census_data_filtered$`Number of households`)
  census_data_filtered$`P_Total` <- as.numeric(census_data_filtered$`P_Total`)
  census_data_filtered$`P_Males` <- as.numeric(census_data_filtered$`P_Males`)
  census_data_filtered$`P_Females` <- as.numeric(census_data_filtered$`P_Females`)
  census_data_filtered$`Area\r\n (In sq. km)` <- as.numeric(census_data_filtered$`Area\r\n (In sq. km)`)
  census_data_filtered$`Population per sq. km.` <- as.numeric(census_data_filtered$`Population per sq. km.`)

  
glimpse(census_data_filtered)  
```

## <span style="font-size: 1.4em;">7. DATA DICTIONARY</span>

```{r warning=FALSE, message=FALSE}
variable_names <- colnames(census_data_filtered)

  library(purrr)
  variable_type <- map_chr(census_data_filtered, class)
  
  description <- c("Official code of each state","Official code of each disctrict","Official code of each sub district", "Classified as State, Union Territory, District, Sub-district, or the entire country - India",
                   "Name of the State, Union Territory, District, Sub-district, or the entire country - India", "Classified as Urban area and Rural area.",
                   "Number of villages inhabited","Number of villages uninhabited","Number of towns in the area", "Number of households in the area","Total population in the area","Total male population in the area",
                   "Total female population in the area", "Total area of the region in square kilometer", "Number of people per square kilometer")
  
  data_dict <- data.frame(variable_names,variable_type,description)
  rownames(census_data_filtered) <- NULL
  data_dict <- data.frame(data_dict, row.names = 1:ncol(census_data_filtered))
  


library(DT)
datatable(data_dict)

```

## <span style="font-size: 1.4em;">8. HANDLING MISSING VALUES</span>

Out of 122 observations having at least 1 NA in all the columns, 117 observations are having valid values in 13 columns. Only 5 are having NAs in all the columns.

```{r warning=FALSE, message=FALSE}

display_df <- census_data_filtered %>% 
    filter_all(any_vars(is.na(.)))
datatable(display_df)

display_df <- census_data_filtered %>% 
    filter_all(all_vars(is.na(.)))
datatable(display_df)

```
## <span style="font-size: 1.15em;">Deleting these 5 rows</span>

```{r warning=FALSE, message=FALSE}

census_data_filtered <- census_data_filtered %>% 
                            filter_all(any_vars(!is.na(.)))

display_df <- census_data_filtered %>% 
    filter_all(all_vars(is.na(.)))
datatable(display_df)

```


## <span style="font-size: 1.4em;">9. Uni-Variate Analysis(After  Imputation)</span>

## <span style="font-size: 1.15em;">Categorical Variable Distribution</span>

```{r warning=FALSE, message=FALSE}
ggplot(data = census_data_filtered, aes(x = `India/ State/ Union Territory/ District/ Sub-district`))+
  geom_bar() 
```


## <span style="font-size: .6em;">To develop a comprehensive strategy for our client, we will filter the census dataset to focus on granular details. Specifically, we will analyze the dataset at the district level to gain deeper insights and provide actionable recommendations.</span>

```{r warning=FALSE, message=FALSE}
census_data_dist <- census_data_filtered %>% 
    filter(`India/ State/ Union Territory/ District/ Sub-district` == 'DISTRICT')
```


## <span style="font-size: 1.15em;">Continuous Variable Distribution</span>

```{r warning=FALSE, message=FALSE}
hist(census_data_dist$P_Total, breaks = 250,
       main = "Total population Distribution", xlim = c(0,5000000))
  boxplot(census_data_dist$`Population per sq. km.`, ylim = c(0,10000))

  plot(density(census_data_dist$`Area\r\n (In sq. km)`, na.rm = TRUE), xlim = c(-2000,30000), main = "Area")
```
  
## <span style="font-size: 1.4em;">10. Imputation</span>
  
## <span style="font-size: 1.15em;">Simple Imputation</span> 

**Area\r\n (In sq. km)**
```{r warning=FALSE, message=FALSE}
census_data_imputed <- census_data_dist
summary(census_data_imputed$`Area\r\n (In sq. km)`)


min_val <- min(census_data_imputed$`Area\r\n (In sq. km)`, na.rm = TRUE)
  max_val <- max(census_data_imputed$`Area\r\n (In sq. km)`, na.rm = TRUE)
  
  imputed_vals <- runif(sum(is.na(census_data_imputed$`Area\r\n (In sq. km)`)), min = min_val, max = max_val)
  
  census_data_imputed$`Area\r\n (In sq. km)`[which(is.na(census_data_imputed$`Area\r\n (In sq. km)`))] <- imputed_vals

  
summary(census_data_imputed$`Area\r\n (In sq. km)`)
```

**Population per sq. km.**

Population per sq. km. can be calculated by dividing the imputed values of Area (In sq. km) by corresponding P_Total

```{r warning=FALSE, message=FALSE}
summary(census_data_imputed$`Population per sq. km.`)

  census_data_imputed$`Population per sq. km.`[which(is.na(census_data_imputed$`Population per sq. km.`))]
  
  
  pd_imputed_vals <-  census_data_imputed$`Area\r\n (In sq. km)`[which(is.na(census_data_imputed$`Population per sq. km.`))]/census_data_imputed$P_Total[which(is.na(census_data_imputed$`Population per sq. km.`))]
  
  
  census_data_imputed$`Population per sq. km.`[which(is.na(census_data_imputed$`Population per sq. km.`))] <- pd_imputed_vals
  
summary(census_data_imputed$`Population per sq. km.`)
```

## <span style="font-size: 1.4em;">10. Uni-Variate Analysis(After  Imputation)</span>

## <span style="font-size: 1.15em;">Comparison</span>

```{r warning=FALSE, message=FALSE}
boxplot(census_data_dist$`Population per sq. km.`,
          main = "Total population Density\n(Before)", ylim = c(0,6000))
  
  boxplot(census_data_imputed$`Population per sq. km.`,
          main = "Total Population Density\n(After)", ylim = c(0,6000))
  
  
  plot(density(census_data_dist$`Area\r\n (In sq. km)`, na.rm = TRUE), xlim = c(-2000,30000), main = "Area\n(Before)")
  plot(density(census_data_imputed$`Area\r\n (In sq. km)`), xlim = c(-2000,30000), main = "Area\n(After)")
```

The before and after comparison shows no significant difference, suggesting good imputation.


## <span style="font-size: 1.4em;">11. Bi-Variate Analysis</span>

```{r warning=FALSE, message=FALSE}

  cor(census_data_imputed[7:15])
  
  plot_correlation(census_data_imputed[7:15])
  
  ggplot(census_data_imputed,aes(x=P_Total,y=NV_Inhabited))+
    geom_point()
```


## <span style="font-size: 1.4em;">12. Final Analysis</span>

Upon examining the Boxplot of Population Density more closely, the presence of upper outliers appears to be a positive sign.

```{r warning=FALSE, message=FALSE}
boxplot(census_data_imputed$`Population per sq. km.`,
          main = "Total Population Density\n(After)", ylim = c(6000,50000))

Recommended_list <- census_data_imputed %>% 
                      filter(`Population per sq. km.` > 6000) %>% 
                      filter(!duplicated(Name)) %>% 
                      pull(Name)

```

Deleting the zones of Delhi and replacing all with 'Delhi'

```{r warning=FALSE, message=FALSE}
Recommended_list <- Recommended_list[c(-9,-10,-11,-12,-13,-14,-15,-16)]  
    
Recommended_list <- c(Recommended_list, "Delhi") 
print(Recommended_list)

phase1 <- c("Chandigarh", "Hyderabad", "Mumbai", "Pune", "Chennai", "Delhi")

phase2 <- c("Ludhiana", "Amritsar", "Karnal", "Panipat", "Faridabad", "Saharanpur", "Muzaffarnagar", 
            "Meerut", "Ghaziabad", "Agra", "Lucknow", "Kanpur Nagar", "Allahabad", "Varanasi", "Patna", 
            "Ahmadabad", "Surat", "Nagpur", "Thane", "Mumbai Suburban", "Bangalore")

phase3 <- c("Kargil", "Kishtwar", "Bijnor", "Moradabad", "Jyotiba Phule Nagar", "Firozabad", 
            "Pilibhit", "Shahjahanpur", "Farrukhabad", "Etawah", "Hamirpur", "Mahoba", "Bahraich", 
            "Shrawasti", "Etah", "Kanshiram Nagar", "Saharsa", "Muzaffarpur", "Siwan", "Vaishali", 
            "Samastipur", "Khagaria", "Bhagalpur", "Munger", "Buxar", "Gaya", "Chandel", "Dhubri", 
            "Dakshin Dinajpur", "Maldah", "North Twenty Four Parganas", "Hugli", "Haora", 
            "Kolkata", "Purbi Singhbhum", "Kendrapara", "Morena", "Datia", "Indore", "Akola", 
            "Koppal", "Ramanagara", "Puducherry","Mahamaya Nagar", "Latur")
```

## <span style="font-size: 1.4em;">13. Recommendation</span>


We have analyzed the list of places with respect to 5G infrastructure and 5G awareness among the public. Based on our analysis, we would recommend a phased approach for rolling out your 5G services in these areas.

The three phases of the rollout would be based on the categorization of the cities as follows:

Phase 1: Cities with the best 5G infrastructure and awareness among the public, including Chandigarh, Hyderabad, Mumbai, Pune, Chennai, and Delhi.

Phase 2: Cities with good 5G infrastructure and awareness among the public, but less than the cities in Phase 1. This includes Ludhiana, Amritsar, Karnal, Panipat, Faridabad, Saharanpur, Muzaffarnagar, Meerut, Ghaziabad, Agra, Lucknow, Kanpur Nagar, Allahabad, Varanasi, Patna, Ahmadabad, Surat, Nagpur, Thane, Mumbai Suburban, and Bangalore.

Phase 3: Cities with average 5G infrastructure and awareness among the public, but less than the cities in Phase 2. This includes cities such as Kargil, Kishtwar, Bijnor, Moradabad, Jyotiba Phule Nagar, Firozabad, Pilibhit, Shahjahanpur, Farrukhabad, Etawah, Hamirpur, Mahoba, Bahraich, Shrawasti, Etah, Kanshiram Nagar, Saharsa, Muzaffarpur, Siwan, Vaishali, Samastipur, Khagaria, Bhagalpur, Munger, Buxar, Gaya, Chandel, Dhubri, Dakshin Dinajpur, Maldah, North Twenty Four Parganas, Hugli, Haora, Kolkata, Purbi Singhbhum, Kendrapara, Morena, Datia, Indore, Akola, Koppal, Ramanagara, and Puducherry.

By following this phased approach, ConnectX can ensure a more efficient rollout of your 5G services, focusing on areas with the highest potential impact in each phase. This will also help the ConnectX to allocate resources more effectively and maximize the return on their investment.

```{r warning=FALSE, message=FALSE}
print(phase1)
print(phase2)
print(phase3)
```
